---
title: "Intro to R"
---

## A Quick Introduction to R

If you have never used R before, this section will give you the foundation you need to follow the rest of this handout. R is a free, open-source programming language widely used for statistical computing and data analysis. Unlike point-and-click software like SPSS, R requires you to type commands—but this makes your work **reproducible**, **transparent**, and **flexible**.

### Installing R and RStudio

You need two things:

1.  **R** — the language itself. Download it from <https://cran.r-project.org/>.
2.  **RStudio** — a user-friendly editor that makes working with R much easier. Download the free version from <https://posit.co/download/rstudio-desktop/>.

Always open **RStudio** (not R directly). RStudio provides a nicer interface with four panels: a script editor, console, environment viewer, and file/plot viewer.

### The RStudio Interface

When you open RStudio, you will see four panes:

-   **Source (top-left):** Where you write and edit your code files.
-   **Console (bottom-left):** Where R actually runs commands. You can type here directly, but it is better to write in the Source pane and run from there.
-   **Environment (top-right):** Shows the data and objects currently in memory.
-   **Files/Plots/Help (bottom-right):** Browse files, view plots, and read help documentation.

### R Basics: Objects and Assignment

In R, you store values in **objects** using the assignment operator `<-`. Think of it as giving a name to a value so you can use it later.

```{r r-basics-objects}
# Store a number
x <- 5
x          # prints 5

# Store text (called a "character string")
greeting <- "Hello, World!"
greeting

# Store a collection of numbers (called a "vector")
scores <- c(1, 2, 3, 4, 3)
scores

# Calculate the mean
mean(scores)

# Store the result in a new object
avg_score <- mean(scores)
avg_score
```

A few things to notice:

-   The `#` symbol starts a **comment**—R ignores everything after `#` on that line.
-   `c()` stands for "combine"—it creates a vector (a list of values).
-   R is **case-sensitive**: `Score`, `score`, and `SCORE` are three different things.

### R Basics: Functions

A **function** takes input, does something with it, and returns output. Functions are called by name followed by parentheses containing the inputs (called **arguments**).

```{r r-basics-functions}
# Some basic functions
mean(scores)      # Average
sd(scores)        # Standard deviation
length(scores)    # How many values?
sum(scores)       # Sum of all values
min(scores)       # Minimum
max(scores)       # Maximum
table(scores)     # Frequency count of each value

# Getting help on any function
?mean           # Opens the help page for mean()
```

### R Basics: Data Frames

Survey data in R is stored in a **data frame**—essentially a table where each column is a variable and each row is an observation. Think of it as a spreadsheet.

```{r r-basics-dataframes}
# Create a small data frame by hand
survey <- data.frame(
  id      = c("A1", "A2", "A3", "A4", "A5"),
  program = c(1, 2, 1, 1, 2),
  score   = c(3, 4, 2, 3, 4)
)

# Print the data frame
survey

# Access a single column with $
survey$program

# Check the dimensions: rows x columns
dim(survey)

# See the structure
str(survey)
```

### R Basics: Installing and Loading Packages

R comes with many built-in functions, but **packages** extend what R can do. A package is a collection of functions that someone else wrote and shared. You need to:

1.  **Install** a package once (downloads it to your computer).
2.  **Load** it every time you start a new R session.

```{r r-basics-packages}
# Install a package (only need to do this ONCE)
# install.packages("tidyverse")

# Load a package (do this EVERY session)
library(tidyverse)
```

Think of installing like buying a book and putting it on your shelf. Loading is like taking it off the shelf and opening it so you can use it.

### R Basics: The Pipe Operator (`%>%`)

One thing you will see throughout this handout is the **pipe operator** `%>%`. It takes the output from one step and passes it as input to the next step. It reads as **"and then."**

```{r r-basics-pipe}
# Without the pipe (nested functions — hard to read):
round(mean(c(3, 4, 2, 3, 4)), 1)

# With the pipe (reads left to right — much clearer):
c(3, 4, 2, 3, 4) %>%
  mean() %>%
  round(1)

# A data example: take the survey data, THEN filter, THEN select columns
survey %>%
  dplyr::filter(program == 2) %>%
  dplyr::select(id, program, score)
```

Reading the piped code out loud: "Start with `survey`, **then** keep only rows where program is 2, **then** select the id, program, and score columns."

### R Basics: Common Pitfalls for Beginners

Here are a few things that trip up new R users:

-   **Forgetting to load a package.** If you get an error like `could not find function "read_excel"`, you probably forgot to run `library(readxl)`.
-   **Typos in variable names.** R is case-sensitive. `Gender` and `gender` are different. Use Tab-completion in RStudio to avoid typos.
-   **Using `=` vs. `<-`.** Both work for assignment, but the convention in R is to use `<-` for assigning values to objects and `=` for setting arguments inside functions.
-   **Not quoting text.** File paths and character values need quotes: `"data.xlsx"`, not `data.xlsx`.

### Setting Your Working Directory

R needs to know where your files are. The **working directory** is the folder R looks in by default.

```{r working-directory}
# Check your current working directory
getwd()

# Set your working directory to where your data file is
# Replace the path below with YOUR actual folder path
# setwd("C:/Users/YourName/Documents/SurveyClass")  # Windows
# setwd("/Users/YourName/Documents/SurveyClass")   # Mac

# Tip: In RStudio, you can also use the menu:
# Session > Set Working Directory > Choose Directory
```

Alternatively, if you use **RStudio Projects** (recommended), your working directory is automatically set to the project folder. To create a project: File \> New Project.

------------------------------------------------------------------------

Now that you have the basics, let's dive into data cleaning!
