---
title: "Statistical Analysis"
---

This handout uses well-validated R packages so students can run core analyses with minimal code. The goal is to focus on interpretation rather than writing long scripts.

How to read this handout:

- Each section starts with a question the analysis answers.
- Each section uses as few lines of R as possible.
- Each section ends with what to report in plain language.

## Packages You Need

Install once if needed:

```{r install-packages, eval=FALSE}
# Run once if needed
install.packages(c(
	"tidyverse", "haven", "janitor", "sjlabelled", "skimr",
	"gtsummary", "rstatix", "GGally", "performance", "report"
))
```

Load at the start of each session:

```{r load-packages}
library(tidyverse)
library(haven)
library(janitor)
library(sjlabelled)
library(skimr)
library(gtsummary)
library(rstatix)
library(GGally)
library(performance)
library(report)
```

## Import SPSS Data

Place `housing.sav` in your project folder. The code below checks several common locations.

```{r import-data}
sav_candidates <- c(
	"housing.sav",
	"data/housing.sav",
	"Data/housing.sav",
	"../housing.sav",
	"~/Downloads/housing.sav"
)

sav_path <- sav_candidates[file.exists(sav_candidates)][1]

if (is.na(sav_path)) {
	message("`housing.sav` not found in common locations. Using a small simulated dataset so this chapter remains runnable.")

	set.seed(5244)
	housing <- tibble(
		price = round(rnorm(200, mean = 350000, sd = 90000), 0),
		sqft = round(rnorm(200, mean = 1900, sd = 500), 0),
		bedrooms = sample(2:5, 200, replace = TRUE),
		bathrooms = sample(1:4, 200, replace = TRUE),
		neighborhood = sample(c("Urban", "Suburban", "Rural"), 200, replace = TRUE),
		renovated = sample(c("No", "Yes"), 200, replace = TRUE)
	)
} else {
	housing <- haven::read_sav(sav_path)
}

housing <- housing %>% janitor::clean_names()

# Convert SPSS labelled variables to factors for easier group analyses.
housing <- housing %>% mutate(across(where(is.labelled), as_factor))

dim(housing)
head(housing)
```

## Step 1: Understand Variables

Question: What variables do we have, and what type is each variable?

Why this matters: Your test choice depends on variable type (numeric vs categorical).

```{r inspect-structure}
glimpse(housing)
skimr::skim(housing)
```

Identify numeric and categorical variables for analysis:

```{r identify-variable-types}
numeric_vars <- housing %>% select(where(is.numeric)) %>% names()
cat_vars <- housing %>% select(where(~ is.factor(.) || is.character(.))) %>% names()

numeric_vars
cat_vars
```

## Step 2: Choose an Outcome Variable

Question: Which variable are we trying to explain or predict?

For housing data, the outcome is often `price`, `value`, or `cost`. The code tries these names first and then defaults to the first numeric variable.

```{r choose-outcome}
target_candidates <- names(housing)[str_detect(names(housing), "price|value|cost|rent")]

outcome_var <- if (length(target_candidates) > 0) {
	target_candidates[1]
} else {
	numeric_vars[1]
}

outcome_var
```

## Step 3: Descriptive Statistics (Low-Code Tables)

Question: What does the sample look like before we run inferential tests?

Interpretation tip:

- Numeric variables: read mean/median/spread.
- Categorical variables: read counts and percentages.

```{r descriptives-table}
housing %>%
  select(all_of(c(outcome_var, head(setdiff(numeric_vars, outcome_var), 4), head(cat_vars, 2)))) %>%
  tbl_summary(missing = "ifany")
```

```{r hist-outcome, fig.width=7, fig.height=4}
ggplot(housing, aes(x = .data[[outcome_var]])) +
	geom_histogram(bins = 30, color = "white", fill = "steelblue") +
	labs(
		title = paste("Distribution of", outcome_var),
		x = outcome_var,
		y = "Count"
	) +
	theme_minimal()
```

## Step 4: Categorical Analysis (Cross-Tab + Chi-Square)

Question: Are two categorical variables associated?

Use this when both variables are categories (for example, `neighborhood` and `renovated`).

```{r choose-two-cat-vars}
chi_candidates <- housing %>%
	select(where(~ is.factor(.) || is.character(.))) %>%
	keep(~ n_distinct(., na.rm = TRUE) >= 2) %>%
	names()

chi_candidates
```

```{r crosstab-and-chi-square}
if (length(chi_candidates) >= 2) {
	cat_var1 <- chi_candidates[1]
	cat_var2 <- chi_candidates[2]

	cat_df <- housing %>%
		select(all_of(c(cat_var1, cat_var2))) %>%
		drop_na() %>%
		mutate(across(everything(), as.factor))

	# Easy-to-read cross-tab with row percentages
	cat_df %>%
		janitor::tabyl(.data[[cat_var1]], .data[[cat_var2]]) %>%
		janitor::adorn_percentages("row") %>%
		janitor::adorn_pct_formatting(digits = 1)

	# Chi-square test of association (robust to character columns)
	chi_table <- table(cat_df[[cat_var1]], cat_df[[cat_var2]])
	chi_table

	stats::chisq.test(chi_table)
} else {
	message("Need at least two categorical variables for cross-tab and chi-square.")
}
```

Bar chart (recommended for categorical comparisons):

```{r bar-chart-categorical, fig.width=7, fig.height=4}
if (exists("cat_var1") && exists("cat_var2")) {
	ggplot(housing, aes(x = .data[[cat_var1]], fill = .data[[cat_var2]])) +
		geom_bar(position = "fill") +
		scale_y_continuous(labels = scales::percent) +
		labs(
			title = paste("Proportion of", cat_var2, "within", cat_var1),
			x = cat_var1,
			y = "Percent",
			fill = cat_var2
		) +
		theme_minimal()
}
```

Pie chart (simple composition view):

```{r pie-chart-categorical, fig.width=6, fig.height=6}
if (exists("cat_var1")) {
	housing %>%
		count(.data[[cat_var1]]) %>%
		ggplot(aes(x = "", y = n, fill = .data[[cat_var1]])) +
		geom_col(width = 1, color = "white") +
		coord_polar(theta = "y") +
		labs(title = paste("Composition of", cat_var1), fill = cat_var1) +
		theme_void()
}
```

## Step 5: Independent-Samples t-test

Question: Is the mean of a numeric outcome different across two groups?

If a categorical variable has exactly two levels (for example, `renovated = Yes/No`), we can compare mean outcome between groups.

```{r choose-binary-group}
binary_candidates <- housing %>%
	select(where(~ is.factor(.) || is.character(.))) %>%
	keep(~ n_distinct(., na.rm = TRUE) == 2) %>%
	names()

binary_candidates
```

```{r ttest-analysis}
if (length(binary_candidates) > 0) {
  group_var <- binary_candidates[1]

  housing %>%
    select(all_of(c(outcome_var, group_var))) %>%
    drop_na() %>%
    t_test(as.formula(paste(outcome_var, "~", group_var)), detailed = TRUE)
} else {
  message("No binary grouping variable found for a t-test.")
}
```

## Step 6: One-Way ANOVA

Question: Is the mean of a numeric outcome different across 3 or more groups?

When a categorical variable has 3 or more groups, compare group means with ANOVA.

```{r choose-anova-group}
anova_candidates <- housing %>%
	select(where(~ is.factor(.) || is.character(.))) %>%
	keep(~ n_distinct(., na.rm = TRUE) >= 3) %>%
	names()

anova_candidates
```

```{r anova-analysis}
if (length(anova_candidates) > 0) {
	anova_var <- anova_candidates[1]

	model_aov <- housing %>%
		select(all_of(c(outcome_var, anova_var))) %>%
		drop_na() %>%
		anova_test(as.formula(paste(outcome_var, "~", anova_var)))

	model_aov

	housing %>%
		select(all_of(c(outcome_var, anova_var))) %>%
		drop_na() %>%
		tukey_hsd(as.formula(paste(outcome_var, "~", anova_var)))
} else {
	message("No 3+ level categorical variable found for ANOVA.")
}
```

## Step 7: Correlation Analysis

Question: How strongly are numeric variables linearly related?

Examine linear relationships among numeric variables.

```{r correlation-matrix}
num_df <- housing %>% select(where(is.numeric))

if (ncol(num_df) >= 2) {
	num_df %>%
		cor_mat() %>%
		cor_get_pval()

	GGally::ggpairs(num_df %>% select(head(names(.), 5)))
} else {
	message("Not enough numeric variables for correlation analysis.")
}
```

## Step 8: Multiple Linear Regression

Question: Which predictors are associated with the outcome when considered together?

Build one model to predict the outcome using up to two numeric and one categorical predictor.

```{r choose-regression-predictors}
predictor_pool <- setdiff(names(housing), outcome_var)

num_predictors <- housing %>%
  select(all_of(predictor_pool)) %>%
  select(where(is.numeric)) %>%
  names() %>%
  head(2)

cat_predictor <- housing %>%
  select(all_of(predictor_pool)) %>%
  select(where(~ is.factor(.) || is.character(.))) %>%
  names() %>%
  head(1)

model_predictors <- c(num_predictors, cat_predictor)
model_predictors
```

```{r regression-fit}
if (length(model_predictors) >= 1) {
  reg_formula <- as.formula(paste(outcome_var, "~", paste(model_predictors, collapse = " + ")))
  reg_fit <- lm(reg_formula, data = housing)

  tbl_regression(reg_fit)
  report(reg_fit)
} else {
  message("No usable predictors found for regression.")
}
```

## Step 9: Regression Assumption Checks

Question: Are key model assumptions reasonable?

Interpretation tip: If diagnostics look poor, report this and consider transformations, robust methods, or simpler models.

```{r regression-diagnostics, fig.width=8, fig.height=8}
if (exists("reg_fit")) {
  check_model(reg_fit)
  check_collinearity(reg_fit)
} else {
  message("Run regression section first to generate diagnostics.")
}
```

## Reporting Template (Simple)

Use this structure to report your findings:

1. Research question and variables.
2. Descriptive table from `tbl_summary()`.
3. Categorical association results from cross-tab and chi-square (`chisq_test()`).
4. Mean-comparison results from `t_test()` or `anova_test()` and post-hoc tests.
5. Regression table from `tbl_regression()` and narrative from `report()`.
6. Practical interpretation in context (for example, what a one-unit change in a predictor means for housing price).
7. Limitations and next steps (missing data, model assumptions, omitted variables).

## Why This Version Is Easier

- Most analyses are one function call.
- Output tables are publication-ready by default.
- Assumption checks are automated with `performance`.
- You can replace auto-selected variables with your planned variables once codebook names are confirmed.
