---
title: "Survey Data Analysis: Weights"
---

## Why do survey weights exist?

Imagine that you want to estimate public opinion in a state.

In the **real population**:

-   80% of people live in urban areas
-   20% of people live in rural areas

But in your **sample**:

-   50% are urban
-   50% are rural

That sample does **not** match the population. If rural and urban respondents answer differently, the raw sample average may be misleading.

In practice, weights are often used because of one or more of these reasons:


- Some people had a higher chance of being sampled than others.
- Some groups were less likely to respond, so their responses may be upweighted.
- The sample is adjusted so that it better matches known population totals, such as age, sex, region, or education.


## A demonstration of survey weights

In this first activity, we will create a population where we know the truth, then draw a biased sample. This helps us see why weights matter.

In the population, we consider two groups: urban (80% population) and rural (20% population). We will make the rural group more supportive of a policy (0.7) than the urban group (0.4).


```{r}
#| warning: false
#| message: false
library(dplyr)
library(ggplot2)
library(survey)

set.seed(123)

N <- 50000

population <- tibble(
  id = 1:N,
  region = sample(c("Urban", "Rural"), N, replace = TRUE, prob = c(0.80, 0.20)),
  supports_policy = rbinom(
    N,
    size = 1,
    prob = ifelse(region == "Urban", 0.4, 0.70)
  )
)

head(population)
```

At the population level, the support rate is ________.

```{r}
true_support <- mean(population$supports_policy)
true_support
```

This value is the **truth** in our simulated population.

Now suppose we draw a sample that is 50% urban and 50% rural. That is a biased sample, because it does not reflect the population composition.

```{r}
set.seed(456)

urban_sample <- population |>
  filter(region == "Urban") |>
  slice_sample(n = 500)

rural_sample <- population |>
  filter(region == "Rural") |>
  slice_sample(n = 500)

sample_data <- bind_rows(urban_sample, rural_sample)

sample_data |> count(region) |> mutate(prop = n / sum(n))
```

::: {.note-box .question}
Will the support rate estimated from this sample be too high, too low, or about right? Run the code below to find out.
:::


```{r}
unweighted_support <- mean(sample_data$supports_policy)
unweighted_support
```


Our sample is 50% urban and 50% rural, but the population is 80% urban and 20% rural.

So we assign weights based on:

$$
	{weight} = \frac{\text{population share}}{\text{sample share}}
$$

```{r}
sample_data <- sample_data |>
  mutate(
    weight = case_when(
      region == "Urban" ~ 0.80 / 0.50,
      region == "Rural" ~ 0.20 / 0.50
    )
  )

sample_data |> count(region, weight)
```



```{r}
weighted_support <- weighted.mean(sample_data$supports_policy, w = sample_data$weight)
weighted_support
```

Now we can compare the three results: the true population value, the unweighted sample estimate, and the weighted sample estimate.

```{r}
comparison <- tibble(
  method = c("True population", "Unweighted sample", "Weighted sample"),
  estimate = c(true_support, unweighted_support, weighted_support)
)

comparison
```


## Analysis using the `survey` package

So far, we used `weighted.mean()`. That is useful for intuition, but survey analysis usually needs methods that also handle **standard errors** appropriately.

The `survey` package is a standard R package for this purpose.

When using `survey` package, we need to specify survey design using `svydesign` function.

- ids = ~1 means that the sampling units are individual rows  and there is no clustering.
- weights = ~weight tells the function to use the specified weights.



```{r}
design_w <- svydesign(
  ids = ~1,
  weights = ~weight,
  data = sample_data
)

design_w
```

To calculate the weighted mean of `supports_policy` with a survey-aware standard error, we can use `svymean()`.
```{r}
svymean(~supports_policy, design = design_w)
```

## An analysis of California schools

The `api` data come from the California Academic Performance Index study included with the `survey` package. The full population data in apipop has 6194 schools.


The key datasets we will use are:

- `apipop`: population of schools
- `apistrat`: stratified sample of schools
- `apiclus1`: one-stage cluster sample of schools
- `apisrs`: simple random sample of schools

Each row represents a school, and `api00` is the school performance score in year 2000. Other useful variables in this handout include:

- `api00`: outcome (school performance)
- `meals`: percent of students eligible for subsidized meals
- `ell`: percent of English language learners
- `stype`: school type (elementary, middle, high)


Let us assume that we want to estimate the average API score in the population of California schools. Because `apipop` is the full population in this teaching dataset, we can compute the population mean directly.

```{r}
data(api)
```

```{r}
true_mean_api <- mean(apipop$api00, na.rm = TRUE)
true_mean_api
```


## Simple random sampling with finite population

```{r}
head(apisrs)
```

When using `survey` package, we need to specify survey design using `svydesign` function.

- ids = ~1 means that the sampling units are individual rows (schools) and there is no clustering.
- fpc = ~fpc tells the function to use the finite population correction based on the `fpc` column in the data, which contains the total population size for each school. 


```{r}
design_srs <- svydesign(
  ids = ~1,
  fpc = ~fpc,
  data = apisrs
)

design_srs
```

::: {.note-box .note}
The `fpc` argument stands for **Finite Population Correction**. Standard variance formulas assume you are sampling from an infinitely large population, which overestimates variance when your sample covers a meaningful fraction of the total population. The FPC adjusts for this by applying the factor $1 - n/N$, where $n$ is the sample size and $N$ is the total population size. In `apisrs`, the `fpc` column holds $N$ (the total number of California schools) for each row. By passing `fpc = ~fpc`, we tell `svydesign()` to use this to shrink standard errors appropriately. If the sampling fraction $n/N$ is negligible (e.g., 200 people from millions), the FPC barely matters — but when it is non-trivial, omitting it meaningfully overstates uncertainty.
:::


To estimate the mean API score (`api00`) with a survey-aware standard error, use `svymean()`.

```{r}
srs_mean <- svymean(~api00, design = design_srs)
srs_mean
```


## Stratified sample analysis

Now let us analyze the stratified sample that comes with the package. The dataset `apistrat` is a stratified sample of schools, where the strata are defined by `stype` (school type).

We specify the survey design for the stratified sample. The `strata` argument tells the function to account for the stratification by school type, and `weights` specifies the sampling weights for each school.

```{r}
strat_design <- svydesign(
  id = ~1,
  strata = ~stype,
  weights = ~pw,
  fpc = ~fpc,
  data = apistrat
)
```

To estimate the mean API score for the stratified sample, we can use `svymean()` again.

```{r}
strat_mean <- svymean(~api00, strat_design)
strat_mean
```


## Clustered sample analysis

Now let us analyze a clustered design.

The dataset `apiclus1` is a one-stage cluster sample. A key cluster identifier is `dnum`, which represents district. This dataset contains 40 districts, and all schools within those districts are included in the sample. The `pw` variable contains the sampling weights for each school.

The code below specifies the survey design for the clustered sample. The `id` argument tells the function that the primary sampling units are districts (`dnum`), and `weights` specifies the sampling weights for each school.

```{r}
clus1_design <- svydesign(
  id = ~dnum,
  weights = ~pw,
  fpc = ~fpc,
  data = apiclus1
)
```

To estimate the mean API score for the clustered sample, we can use `svymean()` again.

```{r}
clus1_mean <- svymean(~api00, clus1_design)
clus1_mean
```



We can now compare three designs:

- simple random sample 
- stratified design
- clustered design

```{r}
results <- tibble(
  design = c("Population truth", "Simple random sample", "Stratified sample", "Clustered sample"),
  estimate = c(
    true_mean_api,
    as.numeric(coef(srs_mean)),
    as.numeric(coef(strat_mean)),
    as.numeric(coef(clus1_mean))
  ),
  se = c(
    NA,
    as.numeric(SE(srs_mean)),
    as.numeric(SE(strat_mean)),
    as.numeric(SE(clus1_mean))
  )
)

results
```




We can also look at confidence intervals for the mean API score under each design. The `confint()` function computes confidence intervals based on the survey design and standard errors.

```{r}
confint(srs_mean)
confint(strat_mean)
confint(clus1_mean)
```

## A regression example

The code below estimates the relationship between API score and percent eligible for subsidized meals using a simple linear regression. We will compare the results from an unweighted model (which ignores the survey design) to a survey-aware model that accounts for clustering and weights.

We will estimate the relationship between API score and percent eligible for subsidized meals.


```{r}
summary(lm(api00 ~ meals, data = apiclus1))

summary(svyglm(api00 ~ meals, design = clus1_design))
```
