---
title: "Reliability Analysis in R"
format:
  html:
    code-copy: true
    code-fold: false
    highlight-style: zenburn
    df-print: paged
    css: ["../assets/sticky-notes.css"]
date: "`r format(Sys.Date(), '%B %d %Y')`"
bibliography: "../assets/epsy8252.bib"
csl: "../assets/apa.csl"
---

```{r}
#| echo: false
source("../assets/notes-setup.R")
```

In this handout, we will discuss how to assess the reliability of survey items and scales in R. 


Recall that reliability refers to the consistency of scores produced by a test.

-   Inconsistency means different scores may be observed for an individual with an unchanged true score when the test was administered more than once.
-   Scores could be inconsistent in varied ways:
    -   consistency of scores over time
    -   consistency of scores over multiple test forms
    -   consistency of scores across test items
    -   consistency of scores across raters
-   Reliability should be assessed accordingly.

## Data

We will use the `bfi` dataset from [here](https://raw.githubusercontent.com/Wenchao-Ma/Data/main/BFI/bfi.csv), which contains item
responses to Big Five personality items on a 1-6 Likert scale to show how to compute reliability in R. The dataset has 25 items, with 5 items for each of the Big Five traits: Extraversion (E), Agreeableness (A), Conscientiousness (C), Neuroticism (N), and Openness (O). To keep the example focused, we will analyze the 5 Agreeableness items (`A1` to
`A5`):

A1
Am indifferent to the feelings of others. (q_146)

A2
Inquire about others' well-being. (q_1162)

A3
Know how to comfort others. (q_1206)

A4
Love children. (q_1364)

A5
Make people feel at ease. (q_1419)

To get the data, use the following code:

```{r load-packages, message=FALSE}

library(psych)
library(tidyr)

# Load built-in dataset
bfi <- read.csv(
  "https://raw.githubusercontent.com/Wenchao-Ma/Data/main/BFI/bfi.csv"
)

# Select Agreeableness items
A_items <- bfi |>
	dplyr::select(A1:A5)

# Keep complete cases for simpler demonstration
A_complete <- A_items |>
	tidyr::drop_na()

dim(A_complete)
head(A_complete)
```


For Likert scale items, make sure item direction is consistent before
computing discrimination, reliability and total score. In this subscale, `A1` is reverse-keyed.

```{r score-items}
# Responses are on a 1-6 scale, so reverse-score with 7 - x
A_keyed <- A_complete |>
	dplyr::mutate(A1 = 7 - A1)

head(A_keyed)
```


## Split-Half Reliability

An early approach to estimating the reliability of a test was to split the test into two halves and correlate the resulting two scores.

::: {.note-box .note}
This approach mimics that of correlating scores on two parallel forms of a test, or on two scores from the same test obtained at different times, but it can be used in situations where neither scores from two test forms nor scores from two test administrations are available.

One acceptable way to split a test is to randomly assign items to one or the other half of the test. Another acceptable way to split a test is to assign odd-numbered items to one half of the test and even-numbered items to the other half.
:::


How to calculate split-half reliability:

-   administer the test to a single sample of individuals
-   The entire set of items is then divided into two equally sized halves, with the intent that the halves are equivalent in terms of item difficulty, item content, and location in the instrument itself (i.e., near the front, in the middle, or near the end)
-   The two halves are then scored individually
-   The correlation between the halves is thus an estimate of the reliability of these **shortened** parallel forms of the measure
-   adjust the correlation between the halves in order to reflect the reliability of the full instrument with the current sample. This can be done via the Spearman-Brown prophecy formula $$\rho_{\text{full test}}=\frac{2\rho_{\text{shortened test}}}{1+\rho_{\text{shortened test}}}$$

In R, we can use the `splitHalf()` function from the `psych` package to compute split-half reliability.
```{r reliability-splithalf}
psych::splitHalf(A_keyed)
```


::: {.note-box .note}
`psych::splitHalf` forms all possible (for n items <= 16) or sample 15,000 (or more) split halfs corrected for test length. This function returns the best and worst splits as item keys that can be used for scoring purposes, if desired. Can do up to 24 items in reasonable time, but gets much slower for more than about 24 items.
:::

## Cronbach’s $\alpha$

Cronbach’s $\alpha$ is probably the most popular internal consistency index, reflecting the degree to which a set of items is interrelated.

The (raw or unstandardized) $\alpha$ coefficient can be calculated by:

$$
\hat{\alpha}=\frac{J}{J-1}\big[1-\frac{\sum_{j=1}^J\hat{\sigma}_{j}^2}{\hat{\sigma}^2}\big]=J^2\frac{\bar{\sigma}_{jj'}}{\hat{\sigma}^2}
$$ 

where $J$ is the test length, $\hat{\sigma}_{k}^2$ the variance of scores on item $j$, and $\hat{\sigma}^2$ the variance of the test scores and $\bar{\sigma}_{jj'}$ the average covariance of all items.

-   Cronbach’s $\alpha$ may be thought of as the mean of all possible split-half correlations, corrected by the Spearman-Brown formula
-   unstandardized $\alpha$ should be used when raw item responses are used and standardized $\alpha$ should be used when standardized item scores are used (Falk, 2011)

In R, we can use the `alpha()` function from the `psych` package to compute Cronbach’s $\alpha$.
```{r reliability-alpha}
psych::alpha(A_keyed)
```

### McDonald’s $\omega$

McDonald’s $\omega$ is another measure for internal consistency, reflecting the degree to which a set of items is interrelated.

Conceptually, McDonald’s $\omega$ is calculated as

$$
\omega=\frac{\text{Signal}}{\text{Signal}+\text{Noise}}.
$$

Computationally, McDonald’s $\omega$ is often calculated based on confirmatory factor analysis.

In R, we can use the `omega()` function from the `psych` package to compute McDonald’s $\omega$.

```{r reliability-alpha-omega}
psych::omega(A_keyed, nfactors = 1)

```

## Recommended Values for internal consistency

Raykov (2011) recommend values of at least .80

Nunnally (1978) recommend values of .70 for scales in the initial stages of development; .80 for “basic research scales,” and .90 as minimum for use in clinical settings



## Exercises

1. Compute the split-half, coefficient alpha and omega reliability for the Extraversion (`E1` to `E5`) and Conscientiousness (`C1` to `C5`) in the `bfi` dataset. Compare the reliability estimates across the two subscales and discuss the results.